---
title: "Math_524_Assignment1"
author: "Saul Hernandez"
date: "2024-09-28"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment 1

**1.1 SVD Results**

We know that $A=UDV^T$ and we are given that;

$A= \begin{bmatrix}1&1\\1&-1 \end{bmatrix}$, $U = \begin{bmatrix} -0.7071068 & -0.7071068\\ -0.7071068& 0.7071068 \end{bmatrix}$, $D = \begin{bmatrix}1.414214 & 0 \\0 & 1.414214\end{bmatrix}$, $V = \begin{bmatrix}-1&0\\1&-1 \end{bmatrix}$

We compute,

$$
\begin{aligned}
UD &= \begin{bmatrix} -0.7071068 & -0.7071068\\ -0.7071068& 0.7071068 \end{bmatrix}\begin{bmatrix}1.414214 & 0 \\0 & 1.414214\end{bmatrix} \\
& = \begin{bmatrix} {-0.7071068 \cdot 1.414214 } & -0.7071068 \cdot 1.414214\\ -0.7071068 \cdot 1.414214 & 0.7071068 \cdot 1.414214 \end{bmatrix} \\
&= \begin{bmatrix}-1&-1\\-1&1 \end{bmatrix}
\end{aligned}
$$ Now we find $(UD)V^T$ as $V^T = V$, we have,

$$
UDV^T = \begin{bmatrix}-1&-1\\-1&1 \end{bmatrix}\begin{bmatrix}-1&0\\0&-1 \end{bmatrix} = \begin{bmatrix}(-1)(-1) + (-1)(0) & (-1)(0)+(-1)(-1)\\(-1)(-1) + (1)(0) & (-1)(0)+(1)(-1) \end{bmatrix} = \begin{bmatrix}1&1\\1&-1 \end{bmatrix}
$$

Thus,

$$
A = \begin{bmatrix}1&1\\1&-1 \end{bmatrix} =  UDV^T 
$$ As desired $\blacksquare$

**1.2 EOF's and PC's**

```{r, 1.2}
library(ggplot2)

setwd("/Users/saulhernandez/Desktop/LinAlg")
darwin_dat = read.table("data/PSTANDdarwin.txt", header = TRUE)
tahiti_dat = read.table("data/PSTANDtahiti.txt", header = TRUE)

head(darwin_dat)
head(tahiti_dat)

#Check dim. & combine
dim(darwin_dat)
dim(tahiti_dat)
comb_dat = cbind(darwin_dat, tahiti_dat)

#SVD
dat_centered = scale(comb_dat)
svd = svd(dat_centered)

# EOFs/PCs
EOFs = svd$u
PCs = svd$d %*% t(svd$v)
```

**1.8 Covariance Matrix; Annual Precipitation**

**(a)** Covariance $N \times N$ matrix choosing an $X$ data matrix from the USHCN annual total precipitation data at three California stations from north to south. $C = X \cdot X^T / Y$ with $N=3$ $Y=5$

```{r, 1.8}
setwd('/Users/saulhernandez/Desktop/LinAlg')
dat = read.csv('CAprcpV2.csv', header = TRUE)
stations =  c('USW00023188', 'USC00047902', 'USW00023234')
years = 2001:2005

filter_dat = subset(dat, STATION %in% stations & DATE %in% years)
prcp_data = filter_dat$PRCP
SDprc = filter_dat$PRCP[filter_dat$STATION == 'USW00023188']
SBprc = filter_dat$PRCP[filter_dat$STATION == 'USC00047902']
SFprc = filter_dat$PRCP[filter_dat$STATION == 'USW00023234']


```

```{r}
#anomaly_mat = rbind(SDprc, SBprc, SFprc)
#rownames(anomaly_mat) =
```

```{r}
#M = colMeans(anomaly_mat)
#A = sweep(anomaly_mat, 2, M)
#A
```

```{r}
#covm = cov(A)
#covm
```

```{r}
#invA = solve (covm)
#eige = eigen(covm)
#V = eige$vectors
#L = eige$values

#SVD = svd(covm)
```

**1.10 Balancing Chemical Reaction Equation**

We can balance the chemical equation $$
\begin{aligned}
{C_{3}H_{8} + O_{2} \rightarrow H_{2}O + CO_{2}}
\end{aligned}
$$ by setting up the system of equations by representing $x_1,x_2,x_3,x_4$ as variables; $$
\begin{aligned}
{x_{1}C_{3}H_{8} + x_{2}O_{2} \rightarrow x_{3}H_{2}O + x_{4}CO_{2}}
\end{aligned}
$$ We see that $$
\begin{aligned}
& C: 3x_1 = x_4 \\
& H: 8x_1 = 2x_3 \\
& O: 2x_2 = 2x_4 + x_3
\end{aligned}
$$ Thus since there are more variables than equations, there are infinitely many solutions. We can set any fixed variable, and express the rest using that variable as such,

$$
\begin{aligned}
x_4 &= 3x_1 \\
x_3 &= 4x_1 \\
x_2 &= 3x_1 + 2x_1 \\
&= 5x_1
\end{aligned}
$$ Thus we the chemical equation is $$
\begin{aligned}
{x_{1}C_{3}H_{8} + 5x_{1}O_{2} \rightarrow 4x_{1}H_{2}O + 3x_{1}CO_{2}}
\end{aligned}
$$ Thus the balanced chemical reaction is $$
\begin{aligned}
C_{3}H_{8} + 5O_{2} \rightarrow 4H_{2}O + 3CO_{2}
\end{aligned}
$$ As desired. $\blacksquare$

**2.1 Generate Space-Time Matrix**

We can generate a Space-Time Matrix using R as such,

```{r,2.1}
setwd("/Users/saulhernandez/Desktop/LinALg")
dat = read.csv("data/NOAAGlobalT.csv", header = TRUE)
#get dimensions
dim(dat)
```

```{r}
dat[1:3, 1:4] #Deleting Rows/Col
```

```{r}
datSD = dat[1777,]
datSD[1444:1551]
```

```{r}
#December
Decind = seq(15, 103, 12)
Decind
```

```{r}
DecSD = datSD[Decind]
row.names(DecSD) = c('San Diego Box')
colnames(DecSD) = 2001:2008
DecSD
```

```{r}
data.frame(DecSD)
```

```{r,2.1.1}
#Another way to try
#SDMat = matrix(datSD[1444:1551], ncol = 12, byrow = TRUE)
#SDMat
#dim(SDMat)
#SDdec = SDMat[,12]
#as.numeric(SDdec)
```

**2.2 Inverse of a matrix** $A = \begin{bmatrix}1.7 & -0.7 & 1.3 \\ -1.6 & -1.4 & 0.4 \\ -1.5 & -0.3 & 0.6 \end{bmatrix}$

```{r,2.2}
A = matrix(c(1.7,-1.6,-1.5,-0.7,-1.4,-0.3,1.3,0.4,0.6), nrow = 3)
A

#Solve inverse
invA = solve (A)
invA 

#Verify
I = invA %*% A 
I
```

**2.3 Linear System of Equations**\
$$
Ax=b
$$

$$
\begin{bmatrix}
1&2&3\\4&5&6\\7&8&9
\end{bmatrix} 
\begin{bmatrix}
x_1 \\ x_2 \\ x_3
\end{bmatrix}
=
\begin{bmatrix}
1 \\ -1 \\ 0
\end{bmatrix}
$$ We can use an R command to find the solution,

```{r, 2.3}

A = matrix(c(1, 2, 3,
              4, 5, 6,
              7, 8, 0), nrow = 3)

b = matrix(c(1, -1, 0))

# Solve for x
x = solve(A, b)

x
```

**2.4 Linear Dependency**

$$
\begin{bmatrix}
1&2&3\\4&5&6\\7&8&9
\end{bmatrix} 
\begin{bmatrix}
x_1 \\ x_2 \\ x_3
\end{bmatrix}
=
\begin{bmatrix}
0 \\ 0 \\ 0
\end{bmatrix}
$$

**(a)** We can show that the row vectors and coefficient matrix are linearly dependent by showing there exists a linear combination; that satisfies $$
c_1x_1 + c_2x_2 + c_3x_3 = 0 
$$

for some constants $c_1,c_2,c_3$ thus, we row reduce

For $R_2 \rightarrow R_2 - 4R_1$:

$$
\begin{bmatrix}
1&2&3 \\ 0&-3&-6 \\ 7&8&9
\end{bmatrix}
$$ Now for $R_3 \rightarrow R_3 - 7R_1$ $$
\begin{bmatrix}
1&2&3 \\ 0&-3&-6 \\ 0&-6&-12
\end{bmatrix}
$$ For $R_3 \rightarrow -\frac{1}{6}R_3$ $$
\begin{bmatrix}
1&2&3 \\ 0&-3&-6 \\ 0&1&2
\end{bmatrix}
$$ For $R_2 \rightarrow R_2 + 3R_3$ $$
\begin{bmatrix}
1&2&3 \\ 0&0&0 \\ 0&1&2
\end{bmatrix}
$$ And Finally, $R_2 \leftrightarrow R_3$ $$
\begin{bmatrix}
1&2&3 \\ 0&1&2 \\ 0&0&0 
\end{bmatrix}
$$ Therefore the zero row implies that the rows are not linearly indeoendent. **(b)** We can reduce the system into two equations by letting $x_3$ be an arbitrary and $x_1,x_2$ are variables;

Using the row reduced matrix from **(a)**, we have the linear system, $$
\begin{aligned}
x_1 + 2x_2 + 3x_3 = 0 \\
x_2 +2x_3 = 0
\end{aligned}
$$ Thus, $x_2=-2x_3$ and we have

$$
\begin{aligned}
x_1 & = -2x_2 - 3x_3 \\
& = -2(-2x_3) - 3x_3 \\
& = 4x_3 - x_3 \\
& = x_3
\end{aligned}
$$ Hence, we have $\begin{bmatrix}x_1 \\ x_2 \\ x_3\end{bmatrix}=x_3\begin{bmatrix}1 \\ -2 \\ 1\end{bmatrix}$

**(c)** $$
x_1 = x_3 \\
x_2 = -2x3
$$

**2.5 Balancing Chemical Reaction Equation**

We can balance the chemical equation $$
\begin{aligned}
{C_{2}H_{6} + O_{2} \rightarrow CO_{2} + H_{2}O}
\end{aligned}
$$ by setting up the system of equations by representing $x,y,z$ as variables; $$
{C_{2}H_{6} + xO_{2} \rightarrow yCO_{2} + zH_{2}O}
$$

We know we have 2 Carbon and 6 Hydrogen in $C_2H_6$, Since $CO_2$ has one Carbon and $H_2O$ has $2$ Hydrogen, we can write $$
\begin{aligned}
2 &= y \\
6 &= 2z\\
z&=3
\end{aligned}
$$ We can find For Carbon, $x$, as such $$
\begin{aligned}
x &= 2y + z\\
& = 2(2) + 3\\
&= 7
\end{aligned}
$$ Thus substituting the molecules back in, the balanced equation is $$
{2C_{2}H_{6} + 7O_{2} \rightarrow 4CO_{2} + 6H_{2}O}
$$

As desired $\blacksquare$

**2.7 Eigenvectors** Show that $u = \begin{bmatrix}1\\1 \end{bmatrix}$ is not an eigenvector of $A=\begin{bmatrix}0&4\\-2&-7 \end{bmatrix}$

**(a)** We can perform Matrix multiplication to show that $u$ is not an eigenvector of $A$, $$
\begin{aligned}
Au =& \begin{bmatrix}0&4\\-2&-7 \end{bmatrix}\begin{bmatrix}1\\1 \end{bmatrix} \\
=& \begin{bmatrix}(0)(1)+(4)(1)\\(-2)(1)+(-7)(1) \end{bmatrix}\\
=& \begin{bmatrix}4\\-9 \end{bmatrix} \neq \lambda u 
\end{aligned} 
$$ For some scalar $\lambda$. Thus $Au \neq \lambda u$. Hence $u$ is not an eigenvector of $A$ $\blacksquare$

**(b)** We can use R to find the unit eigenvectors of matrix $A$

```{r, 2.7}
A = matrix(c(0,-2,4,-7), nrow = 2)
eig = eigen(A)
lambdas = eig$values
Vs = eig$vectors
lambdas
Vs
```

We can check answers by checking if $(A-\lambda I)v = 0$ and $det(A-\lambda I) = 0$ are true,

```{r}
library(geometry)
#check for each value in each vector
for (i in 1:length(lambdas)) { 
  lambda = lambdas[i]
  v = Vs[, i] 
}

I = diag(2)
```

```{r}
{soln = A-(lambda*I)}
soln
```

```{r}
ver = soln %*% v
round(ver)
```

```{r}
det = det(soln)
round(det)
```

As desired $\blacksquare$

\`\`\`\`

\`\`\`\`
